Constraints on Assembly Bias from Galaxy Clustering

### 2016-05-27 ms.tex
Started adding text to subsection:halotools

@@

@@

@@

@@

@@

%--------------------------No AB Triangle Plot-----------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr20_covar_triangle_1.pdf}
\caption{
Two-dimensional marginalized constraints on HOD parameters inferred from 
standard HOD fits to $\wprp(\rp)$ data for the $M_r<-20$ sample. The HOD parameter 
$\log (M_0)$ is extremely poorly constrained by the $\wprp(\rp)$ data and has been 
omitted. The inner contours contain $68\%$ of the posterior probability while the 
outer contours contain $95\%$ of the probability. The panels along the diagonal 
show the one-dimensional, marginalized posteriors on each of these parameters.
}
\label{fig:Mr20triangle}
\end{center}
\end{figure*}
%----------------------------------------------------------------------------------------------------------------------------------

The inferred parameters from our standard analyses differ in several ways from the \citet{zehavi_etal11} 
analysis. Firstly, in our re-analysis of the projected clustering data, we generally find all mass scales to 
be slightly higher than in the work of \citet{zehavi_etal11}. This difference is largely due to the slightly 
different cosmologies adopted in this work. The most important difference are in the values of $\Omega_{\rm M}$, and $\sigma_8$.  
\citet{zehavi_etal11} assumed $\Omega_{\rm M}=0.25$ and $\sigma_8=0.8$, 
whereas in the present work, we use the BolshoiP simulation in which $\Omega_{\rm M}=0.307$ and $\sigma_8=0.82$. 
Larger mass scales are necessary in an analysis with higher $\Omega_{\rm M}$ and $\sigma_8$ in order to maintain galaxy 
number densities with larger halo number densities.

@@

Results of standard HOD fits to SDSS DR7 $w_{\rm p}(r_{\rm p})$ as well as fits using a parameterized model of assembly bias. 

@@

does not include assembly bias.

@@

A second noteworthy difference between the present work and that of \citet{zehavi_etal11} is that we 
find many parameters to be notably more poorly constrained. At the lower luminosity thresholds, for example, 
we constrain $\log (M_{\rm min})$ and $\sigma_{\log M}$ with several times lower precision than 
\citet{zehavi_etal11}. As is evident in Figure~\ref{fig:Mr20triangle}, $\log (M_{\rm min})$ and $\sigma_{\log M}$ 
share a relatively narrow degeneracy. This degeneracy is largely induced by the measured number density of 
the sample. Increasing $\log (M_{\rm min})$ decreases galaxy number density, but this can be compensated by 
an increase in $\sigma_{\log M}$, which places galaxies in a fraction of the considerably more numerous halos 
with masses less than $M_{\rm min}$.

We do not show our constraints on $\log (M_0)$ as they are very poor in all cases, 
with 1-sigma constraints on the order of $\pm \ge 1$~dex for nearly all samples. In several cases, the constraint on 
$\log (M_0)$ is determined by the prior given in Table~\ref{table:priors}. This is in stark contrast to several of the 
results of \citet{zehavi_etal11}. For example, for the threshold sample with $M_r < -19.5$ ($M_r < -20.5$), 
\citet{zehavi_etal11} quote $\log (M_0) = 12.23 \pm 0.17$ ($12.35 \pm 0.24$), whereas we infer 
$\log (M_0) = 11.38^{+0.95}_{-1.57}$ ($11.19^{+0.89}_{-1.39}$). Examining the form of Eq.~(\ref{eq:nsat}), 
it is sensible that the parameter $\log (M_0)$ should be unconstrained at the lower end, because the 
value of $M_0$ does not alter the predicted satellite number once $M_0 \ll M_1$. Therefore, it seems likely 
that the tighter constraints quoted by \citet{zehavi_etal11} must be an error or must be driven by some 
prior that is not clearly specified in their manuscript.

The results of this subsection demonstrate that we achieve reasonable fits to projected galaxy clustering 
data using direct HOD population of a high-resolution numerical simulation of structure formation. These results 
also update existing constraints in the literature in two respects. First, we work within the best-fit Planck cosmology. 
Second, we perform our parameter inference analysis using direct population of halos identified in a numerical 
simulations. This greatly mitigates modeling uncertainties associated with nonlinear density field evolution, 
scale-dependent halo bias, halo exclusion, or other effects that have been difficult to incorporate into 
analytical halo models with high precision.

@@

We turn now to a discussion of our parameter inference analysis of projected galaxy clustering 
in Decorated HOD models that include a treatment of galaxy assembly bias. In this work, 
we consider only the simplest model of galaxy assembly bias, introducing only two new 
parameters, $A_{\rm cen}$ and $A_{\rm sat}$, that describe the strength of central galaxy 
and satellite galaxy assembly bias respectively. These parameters are limited to values 
of $-1 \le A_{\rm cen,sat} \le 1$, and $A_{\rm cen,sat}=0$ when there is no galaxy assembly bias. 
In this work, we use halo concentration as our secondary halo property, so $A_{\rm cen,sat}=1$ 
($A_{\rm cen,sat}=-1$) means that the mean number of galaxies per halo is maximally 
correlated (anti-correlated) with halo concentration. The model and its implementation 
in {\tt halotools} is discussed further in Section~\ref{subsection:decorated} above.

The one-dimensional marginalized constraints on all parameters form these analyses 
are given in the lowest row of each luminosity threshold grouping in Table~\ref{table:parameters}. 
In cases where the posterior on a parameter is monotonic within the physical parameter 
range, we quote an upper or lower limit on the parameter.

Examples of our fits are given in the right-hand panels of Figure~\ref{fig:Mr19samples}, 
Figure~\ref{fig:Mr20samples}, and Figure~\ref{fig:Mr21samples}. The general trend that 
can be gleaned from these figures is that introducing assembly bias improves the ability 
of the predicted two-point functions to match the measured two-point functions across the 
transition form the one-halo (highly nonlinear) to two-halo (nearly linear) regimes near $\rp \sim 2\, h^{-1}{\mathrm{Mpc}}$. 
Visually, these differences appear to be small; however, Table~\ref{table:parameters} shows that 
they are statistically important.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr20ABTri.pdf}
\caption{
The same as Figure~\ref{fig:Mr19ABtriangle}, but for the $M_r<-20$ sample.
}
\label{fig:Mr20ABtriangle}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr21ABTri.pdf}
\caption{
The same as Figure~\ref{fig:Mr19ABtriangle}, but for the $M_r<-21$ sample.
}
\label{fig:Mr21ABtriangle}
\end{center}
\end{figure*}

>>>>

Important points:
-supersedes previous analyses even for standard HOD.
-cannot rule out assembly bias using galaxy clustering.
-some samples favor assembly bias.

@@

For more than a decade, halo occupation modeling has been used to 
interpret large-scale structure measurements and exploit these measurements 
to constrain galaxy formation models and cosmology \citep[e.g.,][]{yang03,tinker05,zehavi05a,
porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
wake_etal11,yang11a,yang12,leauthaud_etal12,tinker_etal13,cacciato_etal13,
more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions underlying 
halo occupation modeling are: (1) all galaxies reside in dark matter halos that are biased 
tracers of the density field; and (2) galaxies occupy halos as a function of halo masses 
only. It is now well known that halo bias depends upon halo properties other than mass
\citep[e.g.][]{gao_etal05,wechsler06,gao_white07,zentner07,dalal_etal08,lacerna11}, 
an effect called halo assembly bias. If galaxies occupy halos as a function of properties 
other than halo mass, then standard halo occupation methods will be subject to a 
systematic error due to galaxy assembly bias. Several of us have previously shown that 
this error can be significant in an analysis of galaxy clustering and can bias inferences 
about many aspects of galaxy evolution \citep{zentner_etal14}. Consequently, we have 
also developed halo occupation models that enable galaxies to occupy halos in a manner 
that depends upon several halo properties \citep{hearin_etal16}. In this paper, we 
revisit the interpretation of luminosity-dependent galaxy clustering in the Sloan Digital Sky 
Survey (SDSS) Data Release 7 (DR7) data, analyzed previously by \citet{zehavi_etal11}, in the context 
both standard halo occupation models and these new models.

Some history here, blah blah blah, Frank loves this stuff. Therefore, I'm sure this part of the introduction 
will get quite lengthy.

Our work is important for several reasons. Our work is a re-analysis of the SDSS DR7 data 
the overcomes shortcomings of previous analyses. In particular, we use direct population of galaxies 
in a cosmological simulation so that delicate issues present in analytic modeling, such as scale-dependent 
halo bias, are treated exactly.  The simulation we use is based on the latest Planck cosmological parameters, 
updating previous work. Furthermore, many important differences between this work and the previous 
work of \citet{zehavi_etal11} were caused by insufficiently large Monte Carlo Markov Chain (MCMC) sampling 
by \citet[][Z. Zheng \& I. Zehavi, Private Communication]{zehavi_etal11}. Therefore, our analysis 
{\em supersedes previous analyses even in the case of standard halo occupation models}.

Furthermore, our work demonstrates explicitly that significant assembly bias in 
$M_r$-selected samples from SDSS DR7 cannot be ruled out based on a standard 
analysis of galaxy clustering only. In fact, we find that several samples {\em 
favor galaxy assembly bias to a degree that is statistically significant}. As demonstrated by 
\citet{zentner_etal14}, this conclusion could have important consequences for the 
interpretation of both extant and forthcoming data.

Our paper is organized as follows. In Section~\ref{section:methods}, we discuss our implementation 
of halo occupation models and our assumptions in our parameter inference analysis. We present 
results for both standard halo occupation analysis and analysis in the context of models with 
galaxy assembly bias in Section~\ref{section:results}. We summarize our results and draw 
conclusions in Section~\ref{section:conclusions}.

@@

To generate predictions for galaxy clustering, we directly populate dark matter halos with mock galaxies using {\tt Halotools}. In this section we review the ``standard" HOD-style model used in the present work, and in the following section we use the Decorated HOD model described in \citet{hearin_etal15}. In both cases, we will only briefly review the salient features of our methodology here; interested readers can always refer to {\tt halotools.readthedocs.io} for comprehensive documentation on all aspects of the {\tt Halotools} framework.

\subsubsection{Occupation statistics}

\subsubsection{Galaxy profiles}

@@

We use the BolshioP simulation in all analyses (give Citation \& Cosmological Parameters).

@@

Short summary of \citet{hearin_etal16}.

@@

%--------------------------No AB Triangle Plot-----------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr20_covar_triangle_1.pdf}
\caption{
Two-dimensional marginalized constraints on HOD parameters inferred from 
standard HOD fits to $\wprp(\rp)$ data for the $M_r<-20$ sample. The HOD parameter 
$\log (M_0)$ is extremely poorly constrained by the $\wprp(\rp)$ data and has been 
omitted. The inner contours contain $68\%$ of the posterior probability while the 
outer contours contain $95\%$ of the probability. The panels along the diagonal 
show the one-dimensional, marginalized posteriors on each of these parameters. 
The values above each panel on the diagonal show the 
median value for the parameter in our chains along with the 
16$^{\rm th}$ and 84$^{\rm th}$ percentiles. 
}
\label{fig:Mr20triangle}
\end{center}
\end{figure*}
%----------------------------------------------------------------------------------------------------------------------------------

The inferred parameters from our standard analyses differ in several ways from the \citet{zehavi_etal11} 
analysis. Firstly, in our re-analysis of the projected clustering data, we generally find all mass scales to 
be slightly higher than in the work of \citet{zehavi_etal11}. This difference is largely due to the slightly 
different cosmologies adopted in this work. The most important difference 
are in the values of $\Omega_{\rm M}$, and $\sigma_8$.  
\citet{zehavi_etal11} assumed $\Omega_{\rm M}=0.25$ and $\sigma_8=0.8$, 
whereas in the present work, we use the BolshoiP simulation in which $\Omega_{\rm M}=0.307$ and $\sigma_8=0.82$. 
Slightly larger mass scales are necessary in an analysis with higher $\Omega_{\rm M}$ and $\sigma_8$ in order 
to maintain galaxy number densities with larger halo number densities.

@@

Results of standard HOD fits to SDSS DR7 $w_{\rm p}(r_{\rm p})$ as well as 
fits using a parameterized model of assembly bias. 

@@

does not include assembly bias. In our analyses, quoted parameter values with errors correspond to the median 
value of the parameter and the 16$^{\rm th}$ and 84$^{\rm th}$ percentiles. In cases for which the posterior 
is monotonic within the physical parameter space, we show one-sided percentiles.

@@

A second noteworthy difference between the present work and that of \citet{zehavi_etal11} is that we 
find many parameters to be notably more poorly constrained. At the lower luminosity thresholds, for example, 
we constrain $\log (M_{\rm min})$ and $\sigma_{\log M}$ with several times lower precision than 
\citet{zehavi_etal11}. We do not show our constraints on $\log (M_0)$ as they are very poor in all cases, 
with 1-sigma constraints on the order of $\pm \ge 1$~dex for nearly all samples. In several cases, the constraint on 
$\log (M_0)$ is determined by the prior given in Table~\ref{table:priors}. This is in stark contrast to several of the 
results of \citet{zehavi_etal11}. For example, for the threshold sample with $M_r < -19.5$ ($M_r < -20.5$), 
\citet{zehavi_etal11} quote $\log (M_0) = 12.23 \pm 0.17$ ($12.35 \pm 0.24$), whereas we infer 
$\log (M_0) = 11.38^{+0.95}_{-1.57}$ ($11.19^{+0.89}_{-1.39}$). Examining the form of Eq.~(\ref{eq:nsat}), 
it is sensible that the parameter $\log (M_0)$ should be unconstrained at the lower end, because the 
value of $M_0$ does not alter the predicted satellite number once $M_0 \ll M_1$. Therefore, it seems likely 
that the tighter constraints quoted by \citet{zehavi_etal11} must be an error.

We have confirmed with a subset of the authors of \citet{zehavi_etal11} 
that the number of MCMC samples they included in their analysis was 
insufficient in a number of cases and that this can lead to a significant underestimation of the uncertainties 
on the inferred parameters, especially $\log (M_{\rm min})$ and $\sigma_{\log M}$ 
(Z. Zheng \& I. Zehavi, Private Communication). The analysis of \citet{zehavi_etal11} 
used $10^4$ samples, whereas we find several $\times 10^6$ samples are often necessary for convergence. 
Additionally, we have recreated qualitatively similar behavior by considering only small subsets of all of our 
full MCMC chains. Consequently, insufficient sampling of the posterior seems to be the likely resolution 
of the discrepancies between our work and that of \citet{zehavi_etal11}.

Two degeneracies are manifest in Fig.~\ref{fig:Mr20triangle} that are common to all of our 
analyses. The parameters $\log (M_1)$ and $\alpha$ are degenerate with each other and positively 
correlated. The parameter $M_1$ is the mass scale at which a halo has one satellite on 
average and $\alpha$ is the power-law index describing the dependence of average satellite 
number on halo mass. Increasing $M_1$ decreases the number of satellites in large halos 
by placing satellites preferentially in halos of larger mass. An increase in $\alpha$ can partly 
compensate for an increase in $M_1$ by increasing the rate at which average satellite number 
grows with halo mass, leading to a larger number of satellites in high-mass halos.

As is evident in Figure~\ref{fig:Mr20triangle}, $\log (M_{\rm min})$ and $\sigma_{\log M}$ 
share a relatively narrow degeneracy as well. This degeneracy is largely induced by the 
measured number density of the sample. Increasing $\log (M_{\rm min})$ decreases galaxy 
number density, but this can be compensated by an increase in $\sigma_{\log M}$, which 
places galaxies in a fraction of the considerably more numerous halos with masses less 
than $M_{\rm min}$. The consequence is that $\log (M_{\rm min})$ and $\sigma_{\log M}$ are 
degenerate with each other such that most of the posterior probability lies in a narrow band 
along which $\log (M_{\rm min})$ and $\sigma_{\log M}$ are positively correlated, as shown in 
Fig.~\ref{fig:Mr20triangle}. In the following plots, we suppress the parameter $\sigma_{\log M}$,  
in order to increase the clarity of the plots, because the viable range of $\sigma_{\log M}$ is 
determined by this simple degeneracy with $\log (M_{\rm min})$.

The results of this subsection demonstrate that we achieve reasonable fits to projected galaxy clustering 
data using direct HOD population of a high-resolution numerical simulation of structure formation. These results 
also update and supersede existing constraints in the literature in at least three respects. First, we work within the 
best-fit Planck cosmology. Second, we perform our parameter inference analysis using direct population of halos 
identified in a numerical simulation of cosmological structure formation (BolshioP). This greatly mitigates modeling 
uncertainties associated with nonlinear density field evolution, scale-dependent halo bias, halo exclusion, 
or other effects that have been difficult to incorporate into analytical halo models with high precision. Third, 
we have explored the posteriors of the parameters with significantly more samples (roughly two orders of magnitude), 
thereby mitigating errors on inferred parameters and their errors induced by insufficient sampling of the 
posterior.

@@

We turn now to a discussion of our parameter inference analysis of projected galaxy clustering 
in Decorated HOD models that include a treatment of galaxy assembly bias. In this work, 
we consider only the simplest model of galaxy assembly bias, introducing only two new 
parameters, $A_{\rm cen}$ and $A_{\rm sat}$, that describe the strength of central galaxy 
and satellite galaxy assembly bias respectively. These parameters are limited to values 
of $-1 \le A_{\rm cen,sat} \le 1$, and $A_{\rm cen,sat}=0$ when there is no galaxy assembly bias. 
In this work, we use halo concentration as our secondary halo property, so $A_{\rm cen,sat}=1$ 
($A_{\rm cen,sat}=-1$) means that the mean number of galaxies per halo is maximally 
correlated (anti-correlated) with halo concentration. The model and its implementation 
in {\tt halotools} is discussed further in Section~\ref{subsection:decorated} above and 
in \citet{hearin_etal16}.

Examples of our fits are given in the right-hand panels of Figure~\ref{fig:Mr19samples}, 
Figure~\ref{fig:Mr20samples}, and Figure~\ref{fig:Mr21samples}. The general trend that 
can be gleaned from these figures is that introducing assembly bias improves the ability 
of the predicted two-point functions to match the measured two-point functions across the 
transition form the one-halo (highly nonlinear) to two-halo (nearly linear) regimes near $\rp \sim 2\, h^{-1}{\mathrm{Mpc}}$. 
This is most apparent for the $M_r < -20$ threshold sample shown in Fig.~\ref{fig:Mr20samples}. 
Visually, these differences appear to be small; however, Table~\ref{table:parameters} shows that 
they are statistically important.

The one-dimensional marginalized constraints on all parameters from these analyses 
are given in the lowest row of each luminosity threshold grouping in Table~\ref{table:parameters}. 
In cases where the posterior on a parameter is monotonic within the physical parameter 
range, we quote an upper or lower limit on the parameter. One- and two-dimensional 
visualizations of the posteriors from our analysis are shown in Figure~\ref{fig:Mr19ABtriangle}, 
Figure~\ref{fig:Mr20ABtriangle}, and Figure~\ref{fig:Mr21ABtriangle}.

@@

%---------------------------------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr20ABTri.pdf}
\caption{
The same as Figure~\ref{fig:Mr19ABtriangle}, but for the $M_r<-20$ sample.
}
\label{fig:Mr20ABtriangle}
\end{center}
\end{figure*}
%---------------------------------------------------------------------------------------------------

%=-------------------------------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr21ABTri.pdf}
\caption{
The same as Figure~\ref{fig:Mr19ABtriangle}, but for the $M_r<-21$ sample.
}
\label{fig:Mr21ABtriangle}
\end{center}
\end{figure*}
%----------------------------------------------------------------------------------------------------

Table~\ref{table:parameters} and Figures~\ref{fig:Mr19ABtriangle}-\ref{fig:Mr21triangle} 
all make several simple, generic points. Introducing additional parameter 
freedom associated with galaxy assembly bias generally increases the viable parameter 
space, even for the subset of standard HOD parameters. This is exactly what is 
expected from the introduction of additional parameter freedom.

Focusing attention on the parameters describing galaxy assembly bias, it is 
evident that these parameters are often quite poorly constrained by galaxy 
clustering data. This implies that galaxy clustering of the precision of SDSS DR7 
measurements cannot rule out, or strongly restrict galaxy assembly bias in most 
cases.

Beyond those generic conclusions, a few specific cases of worthy of further examination. 
Consider the $M_r < -20$ sample. The inferred value of $A_{\rm cen} > 0.28$ at 99\% 
confidence. In this case the data strongly prefer $A_{\rm cen} > 0$ and thus strongly prefer 
galaxies to reside in halos of larger concentration at fixed halo mass.

Changes in Bayesian Information Criterion, for inclusion in some sort of table. Strong evidence 
is usually considered to be a difference $>5$.

Delta BIC 19: 4.37

Delta BIC 19.5: 0.26

Delta BIC 20: 4.56

Delta BIC 20.5: 1.33

Delta BIC 21: -0.54
